The Hyperedge Event Model

### 2018-05-17 paper/ba-HEM/ba-sample.tex
ba-HEM.pdf

Our inference goal is to invert the generative process to obtain the posterior distribution over the unknown parameters, conditioned on the observed data and hyperparamters $(\boldsymbol{\mu}_b, \Sigma_b, \boldsymbol{\mu}_\eta, \Sigma_\eta)$. We draw the samples using Markov chain Monte Carlo (MCMC) methods, repeatedly resampling the value of each parameter from its conditional posterior given the observed data, hyperparamters, and the current values of the other parameters. We express each parameter’s conditional posterior in a closed form using the data augmentation schemes in $\boldsymbol{u}$ \citep{tanner1987calculation}. In this section, we provide each latent variable's conditional posterior.

First, since $u_{adr}$ is a binary random variable, new values may be sampled directly using
\begin{equation}
\begin{aligned}
&P(u_{adr}=1| \boldsymbol{u}_{ad\backslash r}, \boldsymbol{b}, \boldsymbol{x})
\propto \mbox{exp}(\lambda_{adr});\\
&P(u_{adr}=0| \boldsymbol{u}_{ad\backslash r},\boldsymbol{b}, \boldsymbol{x})\propto \text{I}(\lVert\boldsymbol{u}_{ad\backslash r}\rVert_1 > 0 ),
\end{aligned}
\label{eqn:latentreceiver}
\end{equation}
where $I(\cdot)$ is the indicator function that is used to prevent from the instances where a sender chooses zero number of recipients.

@@

	\STATE	Summarize the results with: last chain of $\boldsymbol{b}$, and last chain of $\boldsymbol{\eta}$ (and $\sigma_\tau^2$) 
\label{alg:MCMC}

@@

 	   where $I(\cdot)$ is an indicator function, and the last covariate is the interaction term between outdegree and hyperedge size. As with the nodal (2, 3, 10, 11) and dyadic (4, 5) effects, the triadic (6--9) effects are designed so that their coefficient have a straightforward interpretation. Dyadic statistics ``send" and ``receive" are defined as above such that these covariates measure the number of emails sent from $a$ to $r$ and $r$ to $a$, respectively, within the last 7 days. In the example of triadic statistics, the covariate ``2-send" counts the pairs of emails involving some node $h$ distinct from $a$ and $r$ such that emails from $a$ to $h$ and $h$ to $r$ are both observed within the last 7 days. Other triadic covariates behave similarly, and their interpretations are also analogous, which are illustrated in Figure \ref{figure:netstats}.

@@

	The HEM has many component parts that need to be specified by the user (i.e., the selection of the event timing features $\boldsymbol{y}$, the recipient selection features $\boldsymbol{x}$, and the event time distribution $f$). Many of these components will be specified based on user expertise (e.g., regarding which features would drive recipient selection), but some decisions may require a data-driven approach to model specification. For example, though theoretical considerations may inform the specification of features, subject-matter expertise is unlikely to inform the decision regarding the family of the event time distribution. Furthermore, since different distribution families (and model specifications more generally) may involve different size parameter spaces, any data-driven approach to model comparison must guard against over-fitting the data. In this section we present a general-purpose approach to evaluating the HEM specification using out-of-sample prediction. We illustrate this approach by comparing alternative distributional families for the event timing component of the model. Here, we specifically compare the predictive performance from two distributions---lognormal and exponential. We particularly choose exponential distribution as an alternative to what we used earlier (i.e., lognormal) since exponential is the most commonly specified distribution for time-to-event data which is also used in the stochastic actor oriented models \citep{snijders1996stochastic} as well as their extensions \citep{snijders2007modeling}.

	We evaluate the model's ability to predict edges and timestamps from the Montgomery County email data, conditioned on their ``training" part of the data. To perform the experiment, we separately formed a test split of each three components---sender, recipients, and timestamps---by randomly selecting ``test" data with probability $p=0.1$. Any missing variables were imputed by drawing samples from their conditional posterior distributions, given the observed data, parameter estimates, and current values of test data. We then run inference to update the latent variables given the imputed and observed data. We iterate the two steps---imputation and inference---multiple times to obtain enough number of estimates for ``test" data. Algorithm \ref{alg:PPE} outlines this procedure in detail. We run the experiment and measure the predictive performance of two separate time distributions using 500 predicted samples, which are depicted in Figure \ref{figure:PPEresults}. First, we compare the correct sender probability for missing documents $\{d:a_d=\mbox{NA}\}$, which corresponds to $\pi_{a_{d}}=P(a_{d} = a^{obs}_{d}|\cdot)$ in Algorithm \ref{alg:PPE}. The plot on the left shows that lognormal distribution reaches higher correct sender probability than exponential distribution, although their centers (e.g., medians) are similar. Secondly, we compute similar measure for recipients---correct recipient probability $P(r_{dr} = r^{obs}_{dr})$ in Algorithm \ref{alg:PPE} where $r^{obs}_{dr}$ is either 1 (received) or 0 (not received)---and visually compare the two distributions in logit scale (i.e., $\log(\frac{P(r_{dr} = r^{obs}_{dr}|\cdot)}{1-P(r_{dr} = r^{obs}_{dr})})$). Since the edge generating process (Section \ref{subsec: Tie}) is not directly affected by the choice of timestamp distribution, there is no significant difference between lognormal and exponential in their performance of predicting unknown recipients. Finally, prediction errors for missing timestamps are measured using the median absolute percentage error (MdAPE) \citep{hyndman2006another}, where we take the median of absolute percentage errors $p_{di}$ for $i=1,\ldots,500$ for $d$th document, calculated as follows:
	\begin{algorithm}[!t]
		\spacingset{1}
		\caption{Out-of-Sample Predictions}
		\label{alg:PPE}
			\STATE {\bfseries Input:} data $ \{ (a_d, \boldsymbol{r}_d, t_d)\}_{d=1}^D$, 
			number of new data to generate $R$,
			hyperparameters
			\vskip 0.1in
			\textbf{Test splits:}	
			\STATE draw test authors with $p=0.1$ (out of $D$ authors) 
			\STATE draw test recipients
			with $p=0.1$ (out of $D\times (A-1)$ recipient indicators $\{\{\boldsymbol{r}_{dr}\}_{r\in [A]_{\backslash a_d}}\}_{d=1}^D$)
			\STATE draw test timestamps with $p=0.1$  (out of $D$ timestamps) 
			\STATE set the ``test" data as ``missing" (NA)
			\vskip 0.1in
			\textbf{Imputation and inference:}	
			\STATE initialize the parameters $(\boldsymbol{b}, \boldsymbol{\eta}, \boldsymbol{u}, \sigma^2_\tau)$
			\FOR{$r=1$ {\bfseries to}  $R$}
			\FOR{$d=1$ {\bfseries to}  $D$}
			\IF{$a_d=$  NA}
			\FOR{$a=1$ {\bfseries to} $A$}
			\STATE compute $\pi_{a} $ using $P(a_d= a | \cdot)=f_{\tau}(\tau_{d}; \mu_{a_dd}, \sigma_\tau^2)\times \prod_{a\neq a_d}\big(1-F_{\tau}(\tau_{d}; \mu_{a d}, \sigma_\tau^2) \big)$
			\STATE draw $a_d \sim \mbox{Multinomial}(\pi_a)$
			\ENDIF
			\FOR{$r\in [A]_{\backslash a_d}$}
			\IF{$r_{dr}=$ NA}
			\STATE draw $r_{dr}$ using $P(r_{dr}= 1 | \cdot)$ and $P(r_{dr}= 0| \cdot)$---equation (7)
			\ENDIF
			\IF{$t_d=$ NA}
			\STATE draw $\boldsymbol{\tau}^{new}_d$ from $f_{\tau}(\tau^{new}_{d}; \mu_{a_d d}, \sigma_\tau^2)\times \prod_{a\neq a_d}\big(1-F_{\tau}(\tau_{d}; \mu_{a d}, \sigma_\tau^2) \big)$ via importance sampling (e.g., proposal distribution $g \sim \mbox{halfcauchy}(5)$)
			\ENDIF
			\STATE run inference and update $(\boldsymbol{b}, \boldsymbol{\eta},\boldsymbol{u}, \sigma^2_\tau)$ given the imputed and observed data
			\STATE store the estimates for ``test" data
		\includegraphics[width=1\textwidth]{img/PPEplot-1.png}	
		\caption {Comparison of predictive performance between lognormal and exponential distributions: boxplots for correct sender probability (\textit{left}), distribution of correct recipient probability in logit (\textit{middle}), and boxplots for median absolute percentage error (\textit{right}).}
		\label{figure:PPEresults}
	\begin{equation*}
	p_{di} = \abs*{\frac{\tau^{obs}_d - \tau^{pred}_{di}}{\tau^{obs}_d}},
	\end{equation*}
	and we present boxplots for the MdAPE estimates on the right plot. As expected, we have huge benefit in the performance of timestamp prediction when we choose lognormal distribution over exponential. This difference can be explained by our variance estimate $\hat{\sigma}_\tau^2= 14.093$ in Section \ref{subsec:Result_email}, because exponential distribution omits the extra information on time-increments contained in the large variance estimate. As illustrated above, we can use this out-of-sample prediction task for two usage---1) provide an effective answer to the question ``how are we filling in the missing information of the emails?" and 2) offer one standard way to determine the time-increment distribution in Section \ref{subsec:Time}.

@@

\bibliographystyle{ba}
\bibliography{multicastnetwork}

>>>>

Our inference goal is to invert the generative process to obtain the posterior distribution over the unknown parameters, conditioned on the observed data and hyperparamters $(\boldsymbol{\mu}_b, \Sigma_b, \boldsymbol{\mu}_\eta, \Sigma_\eta)$. We draw the samples using Markov chain Monte Carlo (MCMC) methods, repeatedly resampling the value of each parameter from its conditional posterior given the observed data, hyperparamters, and the current values of the other parameters. We express each parameter’s conditional posterior in a closed form using the data augmentation schemes in $\boldsymbol{u}$ \citep{tanner1987calculation}. In this section, we provide each latent variable's conditional posterior, and demonstrate how we perform the prior-posterior simulator test of \cite{geweke2004getting} for the HEM.

\subsection{Full Conditional Distributions}\label{subsec:conditionaldist}
First, since $u_{adr}$ is a binary random variable, new values may be sampled directly using
\begin{equation}
\begin{aligned}
&P(u_{adr}=1| \boldsymbol{u}_{ad\backslash r}, \boldsymbol{b}, \boldsymbol{x})
\propto \mbox{exp}(\lambda_{adr});\\
&P(u_{adr}=0| \boldsymbol{u}_{ad\backslash r},\boldsymbol{b}, \boldsymbol{x})\propto \text{I}(\lVert\boldsymbol{u}_{ad\backslash r}\rVert_1 > 0 ),
\end{aligned}
\label{eqn:latentreceiver}
\end{equation}
where $I(\cdot)$ is the indicator function that is used to prevent from the instances where a sender chooses zero number of recipients.

@@

	\ENDFOR
	\STATE	Summarize the results with: last chain of $\boldsymbol{b}$, and last chain of $\boldsymbol{\eta}$ (and $\sigma_\tau^2$) 
\end{algorithmic}
\label{alg:MCMC}
\end{algorithm}
\subsection{Getting It Right (GiR) Test} \label{subec:GiR}
  Software development is integral to the objective of applying the HEM to real world data. Code review is a valuable process in any research computing context, and the prevalence of software bugs in statistical software is well documented \citep[e.g., ][]{altman2004numerical,mccullough2009accuracy}.  With highly complex models such as the HEM, there are many ways in which software bugs can be introduced and go unnoticed. As such, we present a joint analysis of the integrity of our generative model, sampling equations, and software implementation.

   \cite{geweke2004getting} introduced the ``Getting it Right'' (GiR) test---a joint distribution test of posterior simulators which can detect errors in sampling equations as well as coding errors---and it has been used to test the implementation of Bayesian inference algorithms \citep{zhao2016bayesian}.  The test involves comparing the distributions of variables simulated from two joint distribution samplers, which we call ``forward" and ``backward" samples. The forward sampler draws unobservable variables from the prior and then generates the observable data conditional on the unobservables. The backward sampler alternates between the inference and an observables simulator, by running the inference code on observable data to obtain posterior estimates of the unobervable variables and then re-generating the observables given the inferred unobservables. The backward sampler is initialized by running an iteration of inference on observables drawn directly from the prior. Since the only information on which both the forward and backward samplers are based is the prior, if the sampling equations are correct and the code is implemented without bugs, each variable should have the same distribution in the forward and backward samples.

   In the forward samples, both observable and unobservable variables are generated using Algorithm \ref{alg:generative}. In the backward samples, unobservable variables are generated using the sampling equations for inference, which we derived in Section \ref{subsec:conditionaldist}. For each forward and backward sample that consists of $D$ number of edges, we save these statistics:
   \begin{itemize}
   	\item[1.] Mean of observed recipient sizes $ \lVert \boldsymbol{r}_{d} \rVert_1 $ across $d=1,\ldots,D$,
   	\item[2.] Mean of time-increments $\tau_d$ across $d=1,...,D$,
   	\item[3.] $b_p$ value used to generate the samples $p = 1,...,P$,
   	\item[4.] $\eta_p$ value used to generate the samples $p = 1,...,P$,
   	\item[5.] $\sigma^2_\tau$ value used to generate the samples
   \end{itemize}

   To keep the computational burden of re-running thousands of rounds of inference manageable, we run the GiR using a relatively small artificial sample, consisting of $D=100$ edges, $A=5$ actors, $P=4$ number of edge covariates, and $Q=3$ number of timestamp covariates per each forward or backward samples, using lognormal distibution for the time distribution $f$. We generated $10^5$ sets of forward and backward samples, and then calculated 1,000 quantiles for each of the statistics. We also calculated t-test and Mann-Whitney test p-values in order to test for differences in the distributions generated in the forward and backward samples. Before we calculated these statistics, we thinned our samples 
   by taking every 9th sample starting at the 10,000th sample for a resulting sample size of 10,000, in order to reduce the autocorrelation in the Markov chains. In each case, if we observe a large p-value, this gives us evidence that the distributions generated under forward and backward sampling have the same locations. We depict the GiR results using probability-probability (PP) plots. To compare two samples with a PP-plot we calculate the empirical quantile in each sample of a set of values observed across the two samples, then plot the sets of quantiles in the two samples against each other. If the two samples are from equivalent distributions, the quantiles should line up on a line with zero $y$-intercept, and unit slope (i.e., a 45-degree line). The GiR test results are depicted in Figure \ref{fig:PPplot}, which show that we pass the test on every statistic.
   \begin{figure}[ht]
   	\centering
   			\includegraphics[width=1\textwidth]{img/etaest-1.png}	
   	\label{fig:PPplot}
   	\caption{Probability-Probability plot for the GiR test statistics.}
   \end{figure}

@@

 	   where $I(\cdot)$ is an indicator function, and the last covariate is the interaction term between outdegree and hyperedge size. As with the nodal (2, 3, 10, 11) and dyadic (4, 5) effects, the triadic (6--9) effects are designed so that their coefficient have a straightforward interpretation. The statistics ``outdegree" and ``indegree" measures the gregariousness and popularity effect of the node by counting the number of emails sent from $a$ and received by $r$, respectively, within the last 7 days. Moreover, in order to capture individual tendency of writing multicast emails, we include the statistic ``hyperedge size"---the number of emails sent from $a$ within last 7 days where emails with $k$ number of recipients are counted as $k$ separate emails---as a variant of outdegree statistic, accounting for hyperedges. Dyadic statistics ``send" and ``receive" are defined as above such that these covariates measure the number of emails sent from $a$ to $r$ and $r$ to $a$, respectively, within the last 7 days. In the example of triadic statistics, the covariate ``2-send" counts the pairs of emails involving some node $h$ distinct from $a$ and $r$ such that emails from $a$ to $h$ and $h$ to $r$ are both observed within the last 7 days. Other triadic covariates behave similarly, and their interpretations are also analogous, which are illustrated in Figure \ref{figure:netstats}.

@@

	The HEM has many component parts that need to be specified by the user (i.e., the selection of the event timing features $\boldsymbol{y}$, the recipient selection features $\boldsymbol{x}$, and the event time distribution $f$). Many of these components will be specified based on user expertise (e.g., regarding which features would drive recipient selection), but some decisions may require a data-driven approach to model specification. For example, though theoretical considerations may inform the specification of features, subject-matter expertise is unlikely to inform the decision regarding the family of the event time distribution. Furthermore, since different distribution families (and model specifications more generally) may involve different size parameter spaces, any data-driven approach to model comparison must guard against over-fitting the data. In this section we present a general-purpose approach to evaluating the HEM specification using out-of-sample prediction. We illustrate this approach by comparing alternative distributional families for the event timing component of the model. Here, we specifically compare the predictive performance from two distributions---lognormal and exponential. We particularly choose exponential distribution as an alternative to what we used earlier (i.e., lognormal) since exponential is the most commonly specified distribution for time-to-event data which is also used in the stochastic actor oriented models \citep{snijders1996stochastic} as well as their extensions \citep{snijders2007modeling}. 
	\begin{algorithm}[!t]
		\spacingset{1}
		\caption{Out-of-Sample Predictions}
		\label{alg:PPE}
		\begin{algorithmic}
			\STATE {\bfseries Input:} data $ \{ (a_d, \boldsymbol{r}_d, t_d)\}_{d=1}^D$, 
			number of new data to generate $R$,
			hyperparameters
			\vskip 0.1in
			\textbf{Test splits:}	
			\STATE draw test authors with $p=0.1$ (out of $D$ authors) 
			\STATE draw test recipients
			with $p=0.1$ (out of $D\times (A-1)$ recipient indicators $\{\{\boldsymbol{r}_{dr}\}_{r\in [A]_{\backslash a_d}}\}_{d=1}^D$)
			\STATE draw test timestamps with $p=0.1$  (out of $D$ timestamps) 
			\STATE set the ``test" data as ``missing" (NA)
			\vskip 0.1in
			\textbf{Imputation and inference:}	
			\STATE initialize the parameters $(\boldsymbol{b}, \boldsymbol{\eta}, \boldsymbol{u}, \sigma^2_\tau)$
			\FOR{$r=1$ {\bfseries to}  $R$}
			\FOR{$d=1$ {\bfseries to}  $D$}
			\IF{$a_d=$  NA}
			\FOR{$a=1$ {\bfseries to} $A$}
			\STATE compute $\pi_{a} $ using $P(a_d= a | \cdot)=f_{\tau}(\tau_{d}; \mu_{a_dd}, \sigma_\tau^2)\times \prod_{a\neq a_d}\big(1-F_{\tau}(\tau_{d}; \mu_{a d}, \sigma_\tau^2) \big)$
			\ENDFOR
			\STATE draw $a_d \sim \mbox{Multinomial}(\pi_a)$
			\ENDIF
			\FOR{$r\in [A]_{\backslash a_d}$}
			\IF{$r_{dr}=$ NA}
			\STATE draw $r_{dr}$ using $P(r_{dr}= 1 | \cdot)$ and $P(r_{dr}= 0| \cdot)$---equation (7)
			\ENDIF
			\ENDFOR
			\IF{$t_d=$ NA}
			\STATE draw $\boldsymbol{\tau}^{new}_d$ from $f_{\tau}(\tau^{new}_{d}; \mu_{a_d d}, \sigma_\tau^2)\times \prod_{a\neq a_d}\big(1-F_{\tau}(\tau_{d}; \mu_{a d}, \sigma_\tau^2) \big)$ via importance sampling (e.g., proposal distribution $g \sim \mbox{halfcauchy}(5)$)
			\ENDIF
			\STATE run inference and update $(\boldsymbol{b}, \boldsymbol{\eta},\boldsymbol{u}, \sigma^2_\tau)$ given the imputed and observed data
			\ENDFOR
			\STATE store the estimates for ``test" data
			\ENDFOR
		\end{algorithmic}
	\end{algorithm}

	We evaluate the model's ability to predict edges and timestamps from the Montgomery County email data, conditioned on their ``training" part of the data. To perform the experiment, we separately formed a test split of each three components---sender, recipients, and timestamps---by randomly selecting ``test" data with probability $p=0.1$. Any missing variables were imputed by drawing samples from their conditional posterior distributions, given the observed data, parameter estimates, and current values of test data. We then run inference to update the latent variables given the imputed and observed data. We iterate the two steps---imputation and inference---multiple times to obtain enough number of estimates for ``test" data. Algorithm \ref{alg:PPE} outlines this procedure in detail. We run the experiment and measure the predictive performance of two separate time distributions using 500 predicted samples, which are depicted in Figure \ref{figure:PPEresults}. First, we compare the correct sender probability for missing documents $\{d:a_d=\mbox{NA}\}$, which corresponds to $\pi_{a_{d}}=P(a_{d} = a^{obs}_{d}|\cdot)$ in Algorithm \ref{alg:PPE}. The plot on the left shows that lognormal distribution reaches higher correct sender probability than exponential distribution, although their centers (e.g., medians) are similar. Secondly, we compute similar measure for recipients---correct recipient probability $P(r_{dr} = r^{obs}_{dr})$ in Algorithm \ref{alg:PPE} where $r^{obs}_{dr}$ is either 1 (received) or 0 (not received)---and visually compare the two distributions in logit scale (i.e., $\log(\frac{P(r_{dr} = r^{obs}_{dr}|\cdot)}{1-P(r_{dr} = r^{obs}_{dr})})$). Since the edge generating process (Section \ref{subsec: Tie}) is not directly affected by the choice of timestamp distribution, there is no significant difference between lognormal and exponential in their performance of predicting unknown recipients. Finally, prediction errors for missing timestamps are measured using the median absolute percentage error (MdAPE) \citep{hyndman2006another}, where we take the median of absolute percentage errors $p_{di}$ for $i=1,\ldots,500$ for $d$th document, calculated as follows:
	\begin{figure}[!t]
		\centering
		\includegraphics[width=1\textwidth]{img/PPEplot-1.png}	
		\caption {Comparison of predictive performance between lognormal and exponential distributions: boxplots for correct sender probability (\textit{left}), distribution of correct recipient probability in logit (\textit{middle}), and boxplots for median absolute percentage error (\textit{right}).}
		\label{figure:PPEresults}
	\end{figure}
	\begin{equation*}
	p_{di} = \abs*{\frac{\tau^{obs}_d - \tau^{pred}_{di}}{\tau^{obs}_d}},
	\end{equation*}
	and we present boxplots for the MdAPE estimates on the right plot. As expected, we have huge benefit in the performance of timestamp prediction when we choose lognormal distribution over exponential. This difference can be explained by our variance estimate $\hat{\sigma}_\tau^2= 14.093$ in Section \ref{subsec:Result_email}, because exponential distribution omits the extra information on time-increments contained in the large variance estimate. As illustrated above, we can use this out-of-sample prediction task for two usage---1) provide an effective answer to the question ``how are we filling in the missing information of the emails?" and 2) offer one standard way to determine the time-increment distribution in Section \ref{subsec:Time}.

@@

\bibliographystyle{ba}
\bibliography{baHEM}
